This is an example of article title

### 2023-07-02 paper/paper.tex
chore: Additions to paper

@@

\section{Theory}
Assuming a set of $n$ $d$-dimensional points $\mathcal{V}$, we represent $\mathcal{V}$ as
the $d \times n$ matrix $V$. \verb|AdaptiveHierarchicalRegularBinning| uses Morton
encoding to map each $d$-dimenensional point of $\mathcal{V}$ to a $1$-dimenensional
space filling curve, Z-order curve.

The mapping proccess requires 3 distinct steps.

\begin{itemize}
  \item Normalize $\mathcal{V}$ such that each vector $\vec{v}$ of $\mathcal{V}$
  satisfies $\vec{v} \in [0,1)^d$
  \item Quantize each dimension using an $l$-bit quantizer
  \item Bit-interleave the output of the quantizer
\end{itemize}

@@

Furthermore $\hat{f}$ must satisfy one more property. Namely, relative distances
must be preserved.

@@

The effect that $\vec{t}$ and $s$ have on the space $\mathcal{H}$ can be computed as such
$$
\begin{matrix}
  \hat{f}(\mathcal{H}) & = & \bigtimes\limits_{i=1}^d [\hat{f}(m_i), \hat{f}(M_i)] &\\
                       & = & \bigtimes\limits_{i=1}^d [\frac{m_i-\vec{t}_i}{s}, \frac{M_i-\vec{t}_i}{s}] &\
\end{matrix}
$$

Defining $\vec{t}$ and $s$ is trivial.

@@

Therefore $\hat{f}(\mathcal{H}) \subseteq [0, 1]^d \Leftrightarrow \hat{f}: \mathcal{H} \rightarrow [0,1]^d$

A subtle, yet important, modification is made to the computed $s$ in order to truely
map each dimension to $[0,1)$ instead of $[0,1]$.

@@

We define $\epsilon(s)$ as the smallest possible value such that $s + \epsilon(s) > s$.
In mathematics when $s \in \mathbb{R}$ there is no such value, due to the uncountably
infinite nature of $\mathbb{R}$. But computers use finite bits to represent their
floating point values and so $\epsilon(s)$ is the machine's epsilon value.

@@

Therefore each dimension is mapped to an $l$-bit word. The bits of
this word approximate the location of the point in the given
dimension.

The output of the quantizer can be mapped $1-1$ to a value in the
original space $[0,1]$. This real value representation can be
trivialy computed as such
$$ \tilde{x}_l = \frac{q^{(l)}(x)}{2^l} $$
Since $2^l-1$ is the max value of the quantizer.

@@

\section{Experiments}
All conducted experiments are executed with a selected radix of $16$ bits and the array
is fully sorted.

\subsection{Dimensionality scaling}
For $n=10^5$ uniformly distributed points we computed the average number of points in a
leaf node of the tree. The graph \ref{b_vs_d} shows that for increasingly higher
dimensional spaces the data get exponatialy sparser.

\begin{figure}[!ht]
  \centerline{
    \includegraphics[width=20pc]{figures/experiments/b_vs_d.png}
  }
  \caption{ Exponential decrease of mean number of points per node }
  \label{b_vs_d}
\end{figure}

\subsection{Runtime performance}
\subsubsection{Scaling with number of points}
For $l=4$ max tree depth and $d=4$ number of dimensions
\verb|AdaptiveHierarchicalRegularBinning.jl| scales linearly with the size of the array
$n$.

@@

\subsubsection{Runtime dimensionality scaling}
High dimensionality should affect the morton encoding since bit-interleaving scales
linearly with the dimensions. The reordering should be affected slightly since each
element

% TODO

\subsubsection{Scaling with number of threads}
For $n=10^6$ points of $d=10$ dimensions and a target tree of $l=4$
levels, we can see from \ref{scale_t} that the scaling of
\verb|AdaptiveHierarchicalRegularBinning.jl| drops exponatialy
with the number of threads.

@@

@@

For uniformly distributed data we expect that each leaf node will contain
$n_{\textrm{node}}$ points.
$$n_{\textrm{node}} = \frac{n}{2^{l\cdot d}}$$

In this case $n = 10^5$, $d = 2$ and $l = 4$. So the expected number of points within a
node is $n_{\textrm{node}} = \frac{10^5}{2^{4\cdot 2}} =390.625$.

\begin{figure}[!ht]
  \centerline{
    \includegraphics[width=20pc]{figures/uniform_2d/hist.png}
  }
  \caption{ Distribution of node length }
  \label{uniform-2d-hist}
\end{figure}

The curse of high dimensionality can be deduced by the formula of $n_{\textrm{node}}$.
The number of possible leaf nodes is $2^{l\cdot d}$ which grows exponentialy with $l$ and
$d$, meaning that for larger $d$ the total number of points $n$ must also grow exponentialy
to keep the ratio constant.

\subsection{Uniform data 10 dimensions}

\begin{lstlisting}[language = Julia]
julia> d, n = size(X)
(10, 100000)

julia> l = 1;

julia> @time t = regural_bin(UInt128,
                             X,
                             l,
                             1;
                             dims=2);
0.046737 seconds (2.48 k allocations: 30.789 MiB)
\end{lstlisting}

\subsection{MNIST data 10 dimensions}
\subsection{PBMC data 10 dimensions}
\subsection{PBMC 36 dimensions}

\section{Introduction}

>>>>

\subsection{Tree}
The resulting tree is memory-dense, meaning that it does not store empty nodes.
The root node encapsulates the entire cloud of points, while each child node points to
a consecutive sub-array of those points. This characteristic holds true for every node
within the tree. Each node is also aware if its level, parent and children.

@@

\section{Theory}
Assuming a set of $n$ $d$-dimensional points $\mathcal{V}$, we represent $\mathcal{V}$ as
the $d \times n$ matrix $V$. \verb|AdaptiveHierarchicalRegularBinning.jl| uses Morton
encoding to map each $d$-dimenensional point of $\mathcal{V}$ to a $1$-dimenensional
space filling curve, Z-order curve.
\\\\
The mapping proccess requires 3 distinct steps.

\begin{itemize}
  \item Normalize $\mathcal{V}$ such that each vector $\vec{v}$ of $\mathcal{V}$
  satisfies $\vec{v} \in [0,1)^d$.
  \item Quantize each dimension using an $l$-bit quantizer, where $l$ is the maximum
  depth of the target tree.
  \item Bit-interleave the output of the quantizer.
\end{itemize}

@@

Furthermore, the function $\hat{f}$ must satisfy an additional property, namely,
the preservation of relative distances.

@@

We require the image of $\hat{f}$ to be $\hat{f}(\mathcal{H}) = [0,1)^d$. Thus we can
compute $\vec{t}$ and $s$ to satisfy that constraint.

$$
\begin{matrix}
  \hat{f}(\mathcal{H}) & \triangleq & \bigtimes\limits_{i=1}^d [\hat{f}(m_i), \hat{f}(M_i)] &\\
                       & = & \bigtimes\limits_{i=1}^d [\frac{m_i-\vec{t}_i}{s}, \frac{M_i-\vec{t}_i}{s}] &\
\end{matrix}
$$

Hence,

@@

Therefore $\hat{f}(\mathcal{H}) \subseteq [0, 1]^d \Leftrightarrow \hat{f}: \mathcal{H} \rightarrow [0,1]^d$
\\\\
A subtle, yet crucial modification is made to the computed $s$ in order to
accurately map each dimension to the interval $[0,1)$, rather than the inclusive range
$[0,1]$.

@@

We define $\epsilon(s)$ as the smallest possible value such that $s + \epsilon(s) > s$.
In the standard real number system, when $s \in \mathbb{R}$, there is no such value,
due to the uncountably infinite nature of $\mathbb{R}$. But computers use finite bits
to represent floating point values and so $\epsilon(s)$ is the machine's epsilon value.

@@

Therefore each dimension is mapped to an $l$-bit word. The bits of
this word approximate the location of the point in the given
dimension.
\\\\
The output of the quantizer can be mapped $1-1$ to a value in the
original space $[0,1)$.

$$ \tilde{x}_l = \frac{q^{(l)}(x)}{2^l} $$

@@

\section{Experiments}
All conducted experiments are executed with a selected radix of $16$ bits and the array
is fully sorted, unless stated otherwise.

\subsection{Curse of high dimensionality}
For $n=10^5$ uniformly distributed points we computed the average number of points in a
leaf node of the tree. The graph \ref{b_vs_d} shows that for increasingly higher
dimensional spaces the data get exponatialy sparser.

\begin{figure}[!ht]
  \centerline{
    \includegraphics[width=20pc]{figures/experiments/b_vs_d.png}
  }
  \caption{ Exponential decrease of mean number of points per leaf node }
  \label{b_vs_d}
\end{figure}

\subsection{Runtime performance}
\subsubsection{Scaling with number of points}
For uniformly distributed $4$-dimensional data and $l=4$ max tree depth, the graph
\ref{scale_n} suggests that \verb|AdaptiveHierarchicalRegularBinning.jl| scales
linearly with the size $n$ of the array.

@@

% TODO
% \subsubsection{Scaling with the number of dimensions}
% High dimensionality should affect the morton encoding since bit-interleaving scales
% linearly with the dimensions. The reordering should be affected slightly since each
% element

\subsubsection{Scaling with number of threads}
For uniformly distributed $n=10^6$ points of $d=10$ dimensions and a target tree of
$l=4$ levels, graph \ref{scale_t} suggests that the runtime of
\verb|AdaptiveHierarchicalRegularBinning.jl| drops exponatialy
with the number of threads.

@@

We would expect that for a large enough number of threads the performance would worsen,
since the runtime complexity of the \verb|Countsort| algorithm suggests that if not
$n \gg t^2 \cdot r$ then the parallel variant is worse than the sequential one. Our
implementation gets around that limitation using the waterfall scheme for its
parallelization and thus for small sub-arrays the sequential variant for the
\verb|Countsort| is used.

@@

For uniformly distributed data we expect that each leaf node will contain
$n_{\textrm{leaf}}$ points.
$$n_{\textrm{leaf}} = \frac{n}{2^{l\cdot d}}$$

In this case $n = 10^5$, $d = 2$ and $l = 4$. So the expected number of points within a
node is $n_{\textrm{leaf}} = \frac{10^5}{2^{4\cdot 2}} =390.625$.

\begin{figure}[!ht]
  \centerline{
    \includegraphics[width=20pc]{figures/uniform_2d/hist.png}
  }
  \caption{ Distribution of leaf node length }
  \label{uniform-2d-hist}
\end{figure}

Graph \ref{uniform-2d-hist} agrees with the theoreticaly expected value, although there
is some variance. This is to be expected, increasing the number of points will reduce
the variance then the histogram will tend more to the expected singular value.

\subsection{Uniform data 10 dimensions}
\subsection{MNIST data 10 dimensions}
\subsection{PBMC data 10 dimensions}
\subsection{PBMC 36 dimensions}

% TODO Remove
\pagebreak
\section{Juliacon stuff, kept for reference}
\section{Introduction}
